A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2018-12-18 article/02_Materials_and_Methods.Rmd
language and grammer

The analysis code of this paper is written in the statistics programming language R [@rcoreteamLanguageEnvironmentStatistical2018] and fully reproducible. A more in-depth description and discussion of the data preparation can be found in the master thesis this article is based on [@schmidComputerbasiertesCulturalEvolution2018].

The dataset used for the analysis in this paper is a subset of the openly available database Radon-B [@kneiselRadonB2013]. It contains ^14^C dates from the European Bronze and Early Iron Age (2300 BC to 500 BC). Radiocarbon dating is one of the major absolute dating techniques for this period [@ramseyRadiocarbonDatingRevolutions2008] and Radon-B is a collection of more than 11000 published dates from around 2800 sites from all over Europe (as of December 2018). Beyond the identifying laboratory number, the actual, uncalibrated date and a literature reference, the ^14^C dates in Radon-B are also documented with context information about the name and location of the dated site (coordinates), the sample material and the the type and cultural attribution of the sampled context. The variable *Feature type* systematically distinguishes for example between settlements, rockshelters and also different kinds of graves: *Grave (cremation), Grave (flat) inhumation*, etc.. This information allows to accurately and precisely determine a moment in space and time, when a grave of a certain type was created. In theory the more radiocarbon dates with this information are available, the better becomes our understanding and reconstruction of burial traditions in prehistoric Europe. Unfortunately the Radon-B dataset has some quality issues mostly due to inconsistent manual data input: Incompleteness and faultiness of meta information, no supervised or well thought-out variable classification for *sample material* and *cultural attribution* and inconsistent multilingualism even within individual entries. These issues though do not inhibit the general data analysis strategy of this paper.

To create a tidy and useful subset of Radon-B, the following major steps of data preparation were applied. Please see the code for details of implementation. The value in brackets indicates the number of dates still available after the respective filtering operation.

@@

10. Transforming data from the meaning *^14^C dates on graves* to *burial type at a certain time*. For `r r("bronze15 multi dates one grave", sf_prep)` entries in the thus far assembled subset the attribution to *site* and *feature* is not unique. That means that for many graves multiple (up to 8) ^14^C dates have been created and entered in Radon-B. Dates can't be treated as semantically equal to graves. There's a variety of phenomena that cause this effect: In some cases safety measurements were taken -- multiple dates for one single grave that all fall within the margin of errors of each other. In other cases there is a set of strongly diverging dates for a collective grave that was used for hundreds of years. There are also cases where the *feature* information in the database is not sufficient to determine whether two dates are really from the same grave despite the identical entry. The following steps were applied to solve the problem automatically. This approach is compatible with the intention to create a cultural trait time series as explained below. (`r r("size bronze17", sf_prep)` graves)

    1. If for an individual ^14^C date no numbers appear in it's *feature* variable value, than it is treated as an individual grave.

    2. For the remaining `r r("bronze16 multi dates one grave with numbers", sf_prep)` problematic cases, groups were created by *site* and *feature*. The cumulated timespan covered by the $2\sigma$ age range of all these dates where taken as the relevant timespan for the respective grave.

11. Defining macro regions to observe the development of *burial type* and *burial construction* within sensible geographical and cultural units that are appropriate to the amount of available data. The regions are circular and were delimited semi-automatically by creating a regular grid in accordance to main cultural-historical regions of the European Bronze Age (**citation**). The regions have a radius of 240km and overlap slightly (Figure \ref{map_regions} A). That means some graves are counted towards not just one but two regions.

12. Reducing the date selection to the space covered by the macro regions and attributing the dates to the respective units. (`r r("gpr size", sf_desc)` graves)

@@

This data for individual graves attributed to a macro region can be transformed to a time series of trait prevalence within the respective regions. If a certain kind of grave can be attributed to a fuzzy point in time with radiocarbon dating, than we also know that most likely a ceremony was held around this time were the respective burial tradition was practised. That means that the cultural trait behind this behaviour must also have been present in one form or the other. An example: The idea *inhumation* must have been present in 1447 calBC in Great Britain since this year lies in the $2\sigma$ age range of the ^14^C date *NZA-32497* which was made for the Amesbury Down inhumation burial *I2639/25217*. For each year in the time frame 2200 -- 800 calBC we can count the amount of graves with dates that cover this year with their $2\sigma$ age range. That allows the construction of region wise time series of absolute occurrence numbers for the variables *burial type* and *burial construction* (Figure \ref{development_burial_type} and \ref{development_burial_construction}). Although the number of dates is very small for some regions and time periods we still can attempt to calculate relative proportions of the mutually exclusive traits (inhumation vs. cremation and mound grave vs. flat grave). That leads to relative time series of trait prevalence, which could be a powerful proxy for the development of the distribution of cultural behaviour.

@@

The continuous time series of year wise burial rite proportions for each region allow to calculate time series of a measure of cultural distance for each region-region relation (Figure \ref{sed_region_matrix}). Several different measures of cultural distance could be used here following the discussion of these matters in the relevant literature (**citation**). In this context the Squared Euclidian Distance (SED) seems to be a sensible statistic, because it is fairly simple, works with only few assumptions and emphasizes differences through squaring.

@@

To test for correlation between the distances based on the variables *burial type* and *burial construction*, the time frame of interest was divided into sections of 200 years. For each section of each the two variables, a matrix of the mean SED over time was calculated. The two matrices of each section can be tested for correlation with the Mantel test, which uses permutations of allegedly correlating matrices to determine how strongly random reordering can alter an underlying correlation coefficient [@mantelDetectionDiseaseClustering1967]. Since both distances compared here, *burial type* and *burial construction* distance, are cardinal scale variables, the Pearson correlation coefficient [@pearsonNotesRegressionInheritance1895] can be applied as a base measure for the Mantel test.

Defining a measure of spatial distance for the above established main units of observation, the macro regions, is only possible in a highly simplified way. A continuous distance measure can't be employed, but it's possible to distinguish ordinal distance classes. For the eight regions four different categories of spatial distance were defined (Figure \ref{map_regions} B). The test for correlation between spatial and cultural distance works similar to the one between the two cultural distances above, except that the Mantel test has to be applied with Spearman's rank correlation coefficient [@spearmanProofMeasurementAssociation1904] here, because of the difference in the data level of measurements. The test result are listed in Table \ref{tab:distanceandcorrelationtable*}.

@@

To get a better understanding how the correlation coefficient should develop if there were strong correlation between spatial and cultural distance as measured by *burial type* and *burial construction*, a simple simulation was implemented based on the model proposed by @neimanStylisticVariationEvolutionary1995. This simulation assumes a population separated into $g$ groups with a group population of $N_g$ each. At the beginning of the simulation there are $k$ ideas randomly, but equally distributed in this population. With every timestep -- $t$ defines the total number of time steps -- all individuals learn from an other randomly chosen individual. The probability that this other individual is from an other group equals $m_i$, for the opposite therefore $1-m_i$. The partner group for each individual learning instance is selected randomly following a group interaction probability defined in the matrix $I$. The distribution of the $k$ variants can be observed for every timestep and documented in a time series like the one for real world data.

>>>>

The analysis code of this paper is written in the statistical programming language R [@rcoreteamLanguageEnvironmentStatistical2018] and fully reproducible. A more in-depth description and discussion of the data preparation can be found in the master thesis this article is based on [@schmidComputerbasiertesCulturalEvolution2018].

The dataset used for the analysis in this paper is a subset of the openly available database Radon-B [@kneiselRadonB2013]. It contains ^14^C dates from the European Bronze and Early Iron Age (2300 BC to 500 BC). Radiocarbon dating is a crucial absolute dating technique for this period [@ramseyRadiocarbonDatingRevolutions2008], and Radon-B is a collection of more than 11000 published dates from around 2800 sites from all over Europe (as of December 2018). Beyond the identifying laboratory number, the actual, uncalibrated date and a literature reference, the ^14^C dates in Radon-B are also documented with context information about the name and location of the dated site (coordinates), the sample material and the type and cultural attribution of the sampled context. The variable *Feature type* systematically distinguishes for example between settlements, rock shelters and also different kinds of graves: *Grave (cremation), Grave (flat) inhumation*, i.a.. This information allows to accurately and precisely determine a moment in space and time when a grave of a particular type was created. In theory, the more radiocarbon dates with this information are available, the better becomes our understanding and reconstruction of burial traditions in prehistoric Europe. Unfortunately, the Radon-B dataset has some quality issues mostly due to inconsistent manual data input: Incompleteness and faultiness of meta information, no supervised or well-thought-out variable classification for *sample material* and *cultural attribution* and inconsistent multilingualism even within individual entries. These issues though do not inhibit the general data analysis strategy of this paper.

To create a tidy and useful subset of Radon-B, the following steps of data preparation were applied. Please see the code for details of implementation. The value in brackets indicates the number of dates still available after the respective filtering operation.

@@

10. Transforming data from the meaning *^14^C dates on graves* to *burial type at a certain time*. For `r r("bronze15 multi dates one grave", sf_prep)` entries in the thus far assembled subset, the attribution to *site* and *feature* is not unique. That means that for many graves multiple (up to 8) ^14^C dates have been created and entered in Radon-B. Dates cannot be treated as semantically equal to graves. There's a variety of phenomena that cause this effect: In some cases, safety measurements were taken -- multiple dates for one single grave that all fall within the margin of errors of each other. In other cases, there is a set of strongly diverging dates for a collective grave that was used for hundreds of years. There are also cases where the *feature* information in the database is not sufficient to determine whether two dates are really from the same grave despite the identical entry. The following steps were applied to solve the problem automatically. This approach is compatible with the intention to create a cultural trait time series as explained below. (`r r("size bronze17", sf_prep)` graves)

    1. If for an individual ^14^C date no numbers appear in its *feature* variable value, then it is treated as an individual grave.

    2. For the remaining `r r("bronze16 multi dates one grave with numbers", sf_prep)` problematic cases, groups were created by *site* and *feature*. The cumulated timespan covered by the $2\sigma$ age range of all these dates were taken as the relevant timespan for the respective grave.

11. Defining macro-regions to observe the development of *burial type* and *burial construction* within sensible geographical and cultural units that are appropriate to the amount of available data. The regions are circular and were delimited semi-automatically by creating a regular grid in accordance to main cultural-historical regions of the European Bronze Age (**citation**). The regions have a radius of 240km and overlap slightly (Figure \ref{map_regions} A). That means some graves are counted towards not just one but two regions.

12. Reducing the date selection to the space covered by the macro-regions and attributing the dates to the respective units. (`r r("gpr size", sf_desc)` graves)

@@

This data for individual graves attributed to a macro region can be transformed to a time series of trait prevalence within the respective regions. If a certain kind of grave can be attributed to a fuzzy point in time with radiocarbon dating, then we also know that most likely a ceremony was held around this time where the particular burial tradition was practised. That means that the cultural trait behind this behaviour must also have been present in one form or the other. An example: The idea *inhumation* must have been present in 1447 calBC in Great Britain since this year lies in the $2\sigma$ age range of the ^14^C date *NZA-32497* which was made for the Amesbury Down inhumation burial *I2639/25217*. For each year in the time frame 2200 -- 800 calBC we can count the number of graves with dates that cover this year with their $2\sigma$ age range. That allows the construction of region wise time series of absolute occurrence numbers for the variables *burial type* and *burial construction* (Figure \ref{development_burial_type} and \ref{development_burial_construction}). Although the number of dates is tiny for some regions and periods, we still can attempt to calculate relative proportions of the mutually exclusive traits (inhumation vs cremation and mound grave vs flat grave). That leads to relative time series of trait prevalence, which could be a powerful proxy for the development of the distribution of cultural behaviour.

@@

The continuous time series of year wise burial rite proportions for each region allow to calculate time series of a measure of cultural distance for each region-region relation (Figure \ref{sed_region_matrix}). Several different measures of cultural distance could be used here following the discussion of these matters in the relevant literature (**citation**). In this context the Squared Euclidian Distance (SED) seems to be a sensible statistic, because it is relatively simple, works with only a few assumptions and emphasises differences through squaring.

@@

To test for correlation between the distances based on the variables *burial type* and *burial construction*, the timeframe of interest was divided into sections of 200 years. For each section of each the two variables, a matrix of the mean SED over time was calculated. The two matrices of each section can be tested for correlation with the Mantel test, which uses permutations of allegedly correlating matrices to determine how strongly random reordering can alter an underlying correlation coefficient [@mantelDetectionDiseaseClustering1967]. Since both distances compared here, *burial type* and *burial construction* distance, are cardinal scale variables, the Pearson correlation coefficient [@pearsonNotesRegressionInheritance1895] can be applied as a base measure for the Mantel test.

Defining a measure of spatial distance for the above established main units of observation, the macro-regions, is only possible in a highly simplified way. A continuous distance measure cannot be employed, but it is possible to distinguish ordinal distance classes. For the eight regions, four different categories of spatial distance were defined (Figure \ref{map_regions} B). The test for correlation between spatial and cultural distance works similar to the one between the two cultural distances above, except that the Mantel test has to be applied with Spearman's rank correlation coefficient [@spearmanProofMeasurementAssociation1904] here, because of the difference in the data level of measurements. The test results are listed in Table \ref{tab:distanceandcorrelationtable*}.

@@

To get a better understanding how the correlation coefficient should develop if there was a strong correlation between spatial and cultural distance as measured by *burial type* and *burial construction*, a simple simulation was implemented based on the model proposed by @neimanStylisticVariationEvolutionary1995. This simulation assumes a population separated into $g$ groups with a group population of $N_g$ each. At the beginning of the simulation, there are $k$ ideas randomly but equally distributed in this population. With every timestep -- $t$ defines the total number of time steps -- all individuals learn from another randomly chosen individual. The probability that this other individual is from another group equals $m_i$, for the opposite therefore $1-m_i$. The partner group for each learning instance is selected randomly following a group interaction probability defined in the matrix $I$. The distribution of the $k$ variants can be observed for every timestep and documented in a time series like the one for real-world data.
